---
title: "R Notebook"
output: html_notebook
---


```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)
library(stringr)
```


# read data
## hospital activity
```{r}
 hospital_ativity_and_patient_demographics <- 
    read_csv(here::here("01_data/hospital_ativity_and_patient_demographics.csv")) %>% 
    clean_names()
  # remove_empty(c("rows", "cols"))
  
  hospital_activity_and_deprivation <- read_csv("01_data/hospital_activity_and_deprivation.csv") %>% 
    clean_names()
```

## admission by covid
```{r}
# by health_board
covid_admission_hb_age_sex <-
  read_csv("01_data/hospitalisations_due_to_covid_19/admissions_by_health_board_age_and_sex.csv") %>% 
  clean_names()

covid_admission_hb_deprivation <- 
  read_csv("01_data/hospitalisations_due_to_covid_19/admissions_by_health_board_and_deprivation.csv") %>% 
  clean_names()
```

##extra data set
```{r}
hospitals <- read_csv(here::here("01_data/healt_board/hospitals.csv")) %>% 
  clean_names()

hb <- read_csv(here::here("01_data/healt_board/health_board.csv")) %>% 
  clean_names()

shb <- read_csv(here::here("01_data/healt_board/special_health_boards.csv")) %>% 
  clean_names()

```


# age_sex -------------------------------------------------------------------

### hospital_ativity_and_patient_demographics
```{r}
activity_patient_demographics <- hospital_ativity_and_patient_demographics %>% 
  mutate(year    = str_sub(quarter, 1, 4), .after = id,
         year    = as.numeric(year),
         quarter = str_sub(quarter, 6), 
         quarter = as.numeric(quarter),
         shb      = if_else(nchar(hb) == 6, hb, NA_character_),
         hb       = if_else(nchar(hb) == 9 & str_detect(hb, '^S08'), hb, NA_character_),
         location = if_else(nchar(location) == 5, location, NA_character_),
         age = str_remove(age, pattern = " years")) %>%
  left_join(x = .,
            y = hb, 
            by = "hb", 
            suffix = c("", "_hb_suffix")) %>% 
  left_join(x = .,
            y = shb, 
            by = "shb", 
            suffix = c("", "_shb_suffix")) %>%
  left_join(x = .,
            y = hospitals, 
            by = "location", 
            suffix = c("", "_hospital_suffix")) %>% 
  select(!ends_with(c("_suffix", "qf"))) %>% 
  relocate(c(20:23, 28:29), .after = 16) %>% 
  select(c(1:22))
```

```{r}
activity_patient_demographics %>% 
  write.csv("../02_cleaned_data/activity_patient_demographics.csv")
```


```{r}

activity_patient_demographics <- read_csv("activity_patient_demographics.csv")


filter(age %in% input$demo_age,
      hb_name %in% input$demo_hb_name,
             admission_type %in% input$demo_admission_type,
             location_name %in% input$demo_location_name)

activity_patient_demographics %>% 
  # filter(sex == "") %>% 
  # filter(age == "20-29") %>% 
  filter(admission_type == "Transfers",
         hb_name == "NHS Fife",
#         location_name == "Golden Jubilee National Hospital",
         age == "10-19") %>% 
  group_by(sex, year, age) %>% 
  summarise(nr_episodes          = sum(episodes), 
             nr_stays             = sum(stays)) %>% 
  #           count_length_episode = sum(length_of_episode),
  #           count_length_stays   = sum(length_of_stay)) %>% 
  ggplot() + 
  aes(x = age, y = nr_episodes, fill = sex) +
   geom_col(position = "dodge") + 
  theme_minimal()

```
```{r}
input <- list(
  demo_age = "10-19",
  demo_hb_name = "NHS Ayrshire and Arran",
  demo_admission_type = "Elective Inpatients"
)

activity_patient_demographics %>% 
  filter(age == input$demo_age,
         hb_name == input$demo_hb_name,
         admission_type == input$demo_admission_type)
```



### hospital_ativity_and_deprivation
```{r}
activity_deprivation <- hospital_activity_and_deprivation %>% 
  mutate(year    = str_sub(quarter, 1, 4), .after = id,
         year    = as.numeric(year),
         quarter = str_sub(quarter, 6), 
         quarter = as.numeric(quarter),
         shb      = if_else(nchar(hb) == 6, hb, NA_character_),
         hb       = if_else(nchar(hb) == 9 & str_detect(hb, '^S08'), hb, NA_character_),
         location = if_else(nchar(location) == 5, location, NA_character_)) %>%  
  left_join(x = .,
            y = hb, 
            by = "hb", 
            suffix = c("", "_hb_suffix")) %>% 
  left_join(x = .,
            y = shb, 
            by = "shb", 
            suffix = c("", "_shb_suffix")) %>%
  left_join(x = .,
            y = hospitals, 
            by = "location", 
            suffix = c("", "_hospital_suffix")) %>% 
  select(!ends_with(c("_suffix", "qf"))) %>% 
  relocate(c(19:22, 27:28), .after = 15) %>% 
  select(c(1:21))
```



```{r}
activity_deprivation %>% 
  write.csv("../02_cleaned_data/activity_deprivation.csv")
```


```{r}
activity_deprivation %>% 
  mutate(simd = factor(simd, levels = c(1, 2, 3, 4, 5))) %>% 
  filter(!is.na(simd)) %>% 
  filter(admission_type == "Transfers") %>% 
  filter(hb_name == "NHS Fife") %>% 
  group_by(year, simd) %>% 
  summarise(nr_episodes = sum(episodes)) %>% 
  ggplot() + 
  aes(x = year, y = nr_episodes, fill = simd) +
  geom_col(position = "dodge") + 
  labs(title = "Activity by Board of Treatment and Deprivation",
       subtitle = "1 - Most deprived | 5 - least deprived",
       x = NULL
       ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2016, 2021, 1)) +
     theme(legend.title = element_blank(),
           legend.position = "bottom",
           panel.grid.minor.x = element_blank(),
           panel.grid.minor.y = element_blank())
```

```{r}
activity_deprivation %>% 
  mutate(simd = factor(simd, levels = c(1, 2, 3, 4, 5))) %>% 
  filter(!is.na(simd)) %>% 
  filter(admission_type == "Elective Inpatients") %>% 
  filter(hb_name == "NHS Ayrshire and Arran") %>% 
  group_by(year, simd, location_name) %>% 
  summarise(nr_episodes = sum(episodes)) %>% 
  slice_max(nr_episodes, n = 10) %>% 
  ggplot() + 
  aes(x = year, y = nr_episodes, fill = simd) +
  geom_col(position = "dodge") + 
  labs(title = "Activity by Board of Treatment and Deprivation",
       subtitle = "1 - Most deprived | 5 - least deprived",
       x = NULL
       ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2016, 2021, 1)) +
     theme(legend.title = element_blank(),
           legend.position = "bottom",
           panel.grid.minor.x = element_blank(),
           panel.grid.minor.y = element_blank()) +
  facet_wrap(~location_name, nrow = 5, ncol = 1)
```

```{r}
activity_deprivation %>%   
mutate(simd = factor(simd, levels = c(1, 2, 3, 4, 5))) %>% 
  filter(!is.na(simd)) %>% 
  filter(admission_type == "Elective Inpatients") %>% 
  filter(hb_name == "NHS Ayrshire and Arran") %>% 
  filter(location_name == "University Hospital Ayr") %>% 
  group_by(year, simd, location_name) %>% 
  summarise(nr_episodes = sum(episodes)) %>% 
  ggplot() + 
  aes(x = year, y = nr_episodes, fill = simd) +
  geom_col(position = "dodge") + 
  labs(title = "Activity by Board of Treatment and Deprivation",
       subtitle = "1 - Most deprived | 5 - least deprived",
       x = NULL
       ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2016, 2021, 1)) +
     theme(legend.title = element_blank(),
           legend.position = "bottom",
           panel.grid.minor.x = element_blank(),
           panel.grid.minor.y = element_blank()) +
  facet_wrap(~location_name)
```




Do people have longer lengths of stay from elective or emergency admissions?
Are there any demographic groups that are driving the activity?
How significant is deprivation as a driver of hospital activity?



## covid_admission_hb_age_sex
```{r}
covid_admission_age_sex <- covid_admission_hb_age_sex %>% 
  mutate(date    = ymd(week_ending), .before = 1,
         year    = year(date),     
         month   = month(date),   
         day     = day(date),     
         quarter = quarter(date),
         hb       = if_else(nchar(hb) == 9 & str_detect(hb, '^S08'), hb, NA_character_)) %>% 
  left_join(x = .,
            y = hb, 
            by = "hb", 
            suffix = c("", "_hb_suffix")) %>% 
  relocate(hb_name, .after = hb) %>% 
  select(!ends_with("qf"), -week_ending) %>% 
  select(c(1:"percent_variation"))
```


```{r}
covid_admission_age_sex %>% 
  write.csv("../02_cleaned_data/covid_admission_age_sex.csv")
```



Do people have longer lengths of stay from elective or emergency admissions?
Are there any demographic groups that are driving the activity?
How significant is deprivation as a driver of hospital activity?



```{r}

```

